---
title: 'Bakery Sales Data Analysis'
author: 'Edward Yu'
date: '2018-11-17'
output:
  html_document:
    number_sections: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: readable
    highlight: tango
---

# Introduction
What we have here is a few months worth of sales data from a bakery. 
The data was pulled from a now deleted Kaggle source, https://www.kaggle.com/deleted-dataset/50231.

We'll start with a little exploratory analysis followed by a Market Basket Analysis.

## Load packages, check data

Let's install the required packages we'll be using.

```{r, message=FALSE,warning=FALSE}
if (!require(tibble)) install.packages("tibble")
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(lubridate)) install.packages("lubridate")
if (!require(ggridges)) install.packages("ggridges")
if (!require(arulesViz)) install.packages("arulesViz")
if (!require(arules)) install.packages("arules")
if (!require(viridis)) install.packages("viridis")
if (!require(ggthemes)) install.packages("ggthemes")
if (!require(gridExtra)) install.packages("gridExtra")
library(ggthemes)
library(gridExtra)
library(tibble)
library(tidyverse)
library(lubridate)
library(ggridges)
library(viridis)
library(arules)
library(arulesViz)
```

And get an idea of what our data holds, and the structure of the data.
```{r,warning=FALSE}
x <- read.csv("data/BreadBasket_DMS.csv") %>%
  mutate(Date=as.Date(Date),
         Time=hms(Time)
         )
summary(x)
glimpse(x)
```

Right off the bat we see our time period is relatively narrow: taken between the dates of 
2016-10-30 and 2017-04-09. In other words, we have only 161 days worth of data.
```{r}
difftime(x$Date[nrow(x)],x$Date[1])
```

We also see most sales happen around 12:45pm and coffee is by far the most popular item.

# Exploration

Let's see what we can glean looking at various manipulations of the data we have available.

## Item frequency

```{r}
x1 <- x %>% 
  group_by(Item) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count)) %>%
  slice(1:10) %>% # keep the top 10 items
  ggplot(aes(x=reorder(Item,Count),y=Count,fill=Item))+
  geom_bar(stat="identity")+
  coord_flip()+
  ggtitle("Most popular line items")+
  theme(legend.position="none")
x1
```

No real surprises with frequency, although the `NONE` is slightly odd.
It should be noted that when using `count` we're adding all instances of that particular item. We are not discriminating between different sales receipts. So if one person has ordered 5 coffees, we count coffee 5 times rather than once.

# Sales over time
We'll explore item sales over various amounts of time.

## Days
Let's get an idea of how many line items are sold throughout the datasets history.
```{r}
# Line items sold per day
x2 <- x %>% 
  group_by(Date) %>% 
  summarise(Count= n()) %>% 
  mutate(Day=wday(Date,label=T)) %>% 
  ggplot(aes(x=Date,y=Count,fill=Day))+
  theme_fivethirtyeight()+
  geom_bar(stat="identity")+
  ggtitle("Line items sold per day")+
  scale_fill_viridis(discrete=TRUE)
x2
```

Plotting total line items sold per day shows some obvious peaks and valleys. Saturday is the big sales day for this bakery, followed by Sunday.

Let's take a closer look.

```{r}
# Total line items sold per weekday
x3 <- x %>% 
  mutate(Day = wday(Date,label=T)) %>% 
  group_by(Day) %>% 
  summarise(Count= n()) %>% 
  ggplot(aes(x=Day,y=Count,fill=Day))+
  theme_fivethirtyeight()+
  geom_bar(stat="identity")+
  ggtitle("Total line items sold per weekday")+
  theme(legend.position="none")
x3
```

We sum all sales ordered by day and confirm the general pattern we saw above: our top sales days are on Saturdays, with Sundays and Fridays following behind. Mondays through Thursdays are all roughly similar in total line items sold.

We've been talking about only line items so far, let's see if anything at all changes when we filter by unique transactions per day rather than line items.

```{r}
# Total unique transactions per weekday
x4 <- x %>% 
  mutate(wday=wday(Date,label=T)) %>% 
  group_by(wday,Transaction) %>% 
  summarise(n_distinct(Transaction)) %>% 
  summarise(Count=n()) %>% 
  ggplot(aes(x=wday,y=Count,fill=wday))+
  theme_fivethirtyeight()+
  geom_bar(stat="identity")+
  ggtitle("Unique transactions per weekday")+
  theme(legend.position="none")
x4
```

In general the trend has remained, with Saturday still at the top with Sunday and Friday close behind. We do have a large difference in scale because not every person buys only one item.

That does bring to mind a good question, how many items to people purchase on average? Let's plot this as we've done above to see if there are any differences throughout the week.
```{r}
x5.1 <- x %>% 
  mutate(Day = wday(Date,label=T)) %>% 
  group_by(Day) %>% 
  summarise(Count= n()) 

x5.2 <- x %>% 
  mutate(wday=wday(Date,label=T)) %>% 
  group_by(wday,Transaction) %>% 
  summarise(n_distinct(Transaction)) %>% 
  summarise(Count=n())

x5.3 <- data.frame(x5.1, # Days, total items
                   x5.2[2], # unique transactions
                   x5.1[2]/x5.2[2])  # items per unique transaction
colnames(x5.3) <- c("Day","Line","Unique","Items.Trans")

ggplot(x5.3,aes(x=Day,y=Items.Trans,fill=Day))+
  theme_fivethirtyeight()+
  geom_bar(stat="identity")+
  ggtitle("Total unique transactions per weekday")+
  theme(legend.position="none")
```

We simply divide total line items by unique transactions to get our items per unique transaction. We can see, in general, two items are the norm with people choosing between 2 and 3 items on Sunday.

Okay let's take a step back and play with line items sold per day. If we filter for top sales days, which days are we most likely to get?

```{r}
# Top 50 sales days per weekday
x6 <- x %>% 
  group_by(Date) %>% 
  summarise(Count= n()) %>% 
  arrange(desc(Count)) %>% 
  slice(1:50) %>% 
  mutate(wday = wday(Date,label=T)) %>% 
  ggplot(aes(x=wday,y=Count,fill=wday))+
  theme_fivethirtyeight()+
  geom_bar(stat="identity")+
  ggtitle("Top 50 sales days by weekday")+
  theme(legend.position="none")
x6
```

It becomes much more clear that Saturdays and Sundays are the busiest days, with Tuesdays and Wednesdays as the slow days.

Let's take an even more detailed look by zooming into the time scale.

## Hours
We do a simple line items sold per hour plot and unsurprisingly see that most items are sold around lunch time.
```{r}
# Line items sold per Hour
x7 <- x %>%
  mutate(Hour = as.factor(hour(x$Time))) %>% 
  group_by(Hour) %>% 
  summarise(Count=n()) %>% 
  ggplot(aes(x=Hour,y=Count,fill=Hour))+
  theme_fivethirtyeight()+
  geom_bar(stat="identity")+
  ggtitle("Line items sold per Hour")+
  theme(legend.position="none")
x7
```

Can we learn anything about unique transactions, as we did above with the date data?
```{r}
# Unique transactions per Hour
x8 <- x %>% 
  mutate(Hour = as.factor(hour(Time)))%>% 
  group_by(Hour,Transaction) %>% 
  summarise(n_distinct(Transaction)) %>% 
  summarise(Count=n()) %>% 
  ggplot(aes(x=Hour,y=Count,fill=Hour))+
  theme_fivethirtyeight()+
  geom_bar(stat="identity")+
  ggtitle("Unique transactions per Hour")+
  theme(legend.position="none")
x8
```

We really don't see much of a differece beside scale. Again, let's see how many items are bought in each transaction.
```{r}
x9.1 <- x %>% 
  mutate(Hour = as.factor(hour(Time)))%>% 
  group_by(Hour) %>% 
  summarise(Count= n()) 

x9.2 <- x %>% 
  mutate(Hour = as.factor(hour(Time)))%>% 
  group_by(Hour,Transaction) %>% 
  summarise(n_distinct(Transaction)) %>% 
  summarise(Count=n())

x9.3 <- data.frame(x9.1, # Days, total items
                   x9.2[2], # unique transactions
                   x9.1[2]/x9.2[2])  # items per unique transaction
colnames(x9.3) <- c("Hour","Line","Unique","Items.Trans")

x9 <- 
  ggplot(x9.3,aes(x=Hour,y=Items.Trans,fill=Hour))+
  theme_fivethirtyeight()+
  geom_bar(stat="identity")+
  ggtitle("Total items per transaction per hour")+
  theme(legend.position="none")+
  geom_text(aes(label=round(Items.Trans,1)), vjust=2)
x9
```

People buy the most items per transaction between 10am and 5pm. Let's see if we can visualize transaction density per hour for all days of the week using a ridge plot
```{r}
x10 <- x %>% 
  mutate(Hour=as.factor(hour(Time)),
         Day=wday(Date,label=T)) %>% 
  group_by(Day,Hour,Transaction) %>% 
  summarise(Count=n()) %>% 
  ggplot(aes(x=Hour,y=Day,group=Day,fill=..density..))+
  theme_fivethirtyeight()+
  ggtitle("Transaction density per Hour by Day")+
  theme(legend.position="none")+
  geom_density_ridges_gradient(scale=2, rel_min_height=0.01)+
  scale_fill_viridis(option="magma")
x10
```

The ridge plot gives a nice summary of what we've seen in the above bar charts. Sunday and Saturday have the most transactions and show the most dense transaction peaks around 11pm. 

It's time for a new take on the data using Market Basket Analysis.

# Market Basket Analysis

Let's import our transaction data to the appropriate format and check it out.
```{r}
y <- read.transactions("data/BreadBasket_DMS.csv",
                       format="single",
                       cols=c(3,4),
                       sep=","
                       )
summary(y)
```

We see our most frequent items agree with what we found above. And that most transactions are either one or two items, including 3 lucky people who apparently bought for the entire office.

If we want to visualize the frequent items using `arules` we can.
We need to import our transaction data, and we can go ahead and reaffirm our previous findings by using the `itemFrequencyPlot` function to find the top 5 most popular items
```{r}
par(mfrow=c(1,2))
itemFrequencyPlot(y,topN=5,type="relative")
itemFrequencyPlot(y,topN=5,type="absolute")
```

Now before we perform our MBA we need to determine reasonable values for the support and confidence values. 

__Support:__ An itemset with high support means it appears in our transactions frequently. High confidence is direclty related to the frequency an item appears. For example, coffee probably has very high support.

__Confidence:_*__ Is the strength of our association rule. For example, a confidence of 1 implies that when the LHS item is purchased, the RHS item is purchased 100% of the time.

So let's pin down our `support` value first. 
```{r}
summary(itemFrequency(y))
```

Remember, item frequency is directly related to support. We can see the distribution of values above, with the max frequency at 0.4820079. This is probably coffee which we saw above appears greater than 40% of the time in purchases. 

Since our maximum support is 48%, let's pick a value around 30% to start. We can find this value by dividing 30 by the total amount of transactions.
```{r}
30/length(y)
```

Our support value will be roughly 0.0045. So what about confidence? Let's use a cointoss as inspiration. Remember that confidence is the likelihood, having purchased the item on the LHS, that the item on the RHS will be purchased in addition. Let's go with 50%, or a value of 0.5.

```{r}
rules <- apriori(y,parameter=list(support=0.0045,confidence=.5))
rules
```

We're give a set of 20 rules, which is quite managable.


